Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-05-02 WholePlotTrajectories.Rmd
Discussions
Table

In tropical forests communities are shaped by a constant range of disturbance that changes the abiotic environment, as the light, heat and water fluxes, and the biotic interaction and competitive pressure.
The cornerstone of tropical forests ecology is then to understand their response to disturbance and the corresponding mechanisms [@White2001;@Chazdon2003a].
Forests response to disturbance has been largely studied through structural parameters rapid and convenient to measure, as aboveground biomass, tree height or stem density.
These structural parameters have then been sucessfully modeled and assessed the maintenance of ecosystems processes and services [@Denslow2000;@Blanc2009;@Rutishauser2016].
However the response of tree species diversity although it is determinant of ecosystems productivity, stability and functioning [@Tilman2014] and it would be most probably impacted by the biotic and abiotic changes after disturbance [@CazzollaGatti2014].

@@

```{r Tab1, echo=FALSE}
Intervention <- data.frame(
  Treatment=c("Control", "T1", "T2", "T3"),
  Timber=c("", 
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10 trees/ha", 
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10 trees/ha",
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10 trees/ha"),
  Thinning=c("", 
             "", 
             "DBH $\\geq$ 40 cm, non-valuable species, $\\approx$ 30 trees/ha", 
             "DBH $\\geq$ 50 cm, non-valuable species, $\\approx$ 15 trees/ha  "),
  Fuelwood=c("", 
             "", 
             "", 
             "40 cm $\\leq$ DBH $\\leq$ 50 cm, non-valuable species, $\\approx$ 15 trees/ha"),
  AGBLost=c(0, "$[12\\%-33\\%]$", "$[33\\%-56\\%]$", "$[35\\%-56\\%]$")
)
names(Intervention)[5] <- "\\%AGB lost"
knitr::kable(Intervention, caption="Intervention table, summary of the disturbance intensity for the 4 plot treatments in Paracou.", longtable = FALSE, booktabs = TRUE, escape = FALSE) %>%
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = TRUE)
```

>>>>

In tropical forests communities are shaped by a constant range of disturbance that changes the abiotic environment, as the light, heat and water fluxes, and the biotic interaction and competitive pressure.
The cornerstone of tropical forests ecology is then to understand their response to disturbance and its mechanisms [@White2001;@Chazdon2003a].
Forests response to disturbance has been largely studied through structural parameters rapid and convenient to measure, as aboveground biomass, tree height or stem density.
These structural parameters have then been sucessfully modeled and assessed the maintenance of ecosystems processes and services [@Denslow2000;@Blanc2009;@Rutishauser2016].
However the response of tree species diversity although it is determinant of ecosystems productivity, stability and functioning [@Tilman2014] and it would be most probably impacted by the biotic and abiotic changes after disturbance [@CazzollaGatti2014].

@@

```{r Tab1, echo=FALSE,is.as=T}
Intervention <- data.frame(
  Treatment=c("Control", "T1", "T2", "T3"),
  Timber=c("", 
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10 trees/ha", 
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10 trees/ha",
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10 trees/ha"),
  Thinning=c("", 
             "", 
             "DBH $\\geq$ 40 cm, non-valuable species, $\\approx$ 30 trees/ha", 
             "DBH $\\geq$ 50 cm, non-valuable species, $\\approx$ 15 trees/ha  "),
  Fuelwood=c("", 
             "", 
             "", 
             "40 cm $\\leq$ DBH $\\leq$ 50 cm, non-valuable species, $\\approx$ 15 trees/ha"),
  AGBLost=c(0, "$[12\\%-33\\%]$", "$[33\\%-56\\%]$", "$[35\\%-56\\%]$")
)
names(Intervention)[5] <- "\\%AGB lost"
knitr::kable(Intervention, caption="Intervention table, summary of the disturbance intensity for the 4 plot treatments in Paracou.", longtable = FALSE, booktabs = TRUE, escape = FALSE) %>%
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = TRUE)
```

### 2018-05-02 WholePlotTrajectories.tex
FunctionalGraph
Table

In tropical forests communities are shaped by a constant range of
disturbance that changes the abiotic environment, as the light, heat and
water fluxes, and the biotic interaction and competitive pressure. The
cornerstone of tropical forests ecology is then to understand their
response to disturbance and the corresponding mechanisms
\citep{White2001, Chazdon2003a}. Forests response to disturbance has
been largely studied through structural parameters rapid and convenient
to measure, as aboveground biomass, tree height or stem density. These
structural parameters have then been sucessfully modeled and assessed
the maintenance of ecosystems processes and services
\citep{Denslow2000, Blanc2009, Rutishauser2016}. However the response of
tree species diversity although it is determinant of ecosystems
productivity, stability and functioning \citep{Tilman2014} and it would
be most probably impacted by the biotic and abiotic changes after
disturbance \citep{CazzollaGatti2014}.

@@

The taxonomic diversity was assessed through Richness and the Hill
number translation of Shannon and Simpson indices \citep{Hill1973}.
\textgreater{}\textgreater{}To tackle the unequal number of recruited
trees among treatments the indices bias corrected estimator were used,
following \citep{Chao2015, Marcon2015b}.

@@

@@

\caption{Trajectories of the plots in terms of \textbf{(a)} flora composition and \textbf{(b)} functional composition regarding the 6 leaf and stem functional traits,the maximum allometric height and seed mass class in the two-dimensional space from the NMDS performed for the 30 years after disturbance. Distance matrix for NMDS were computed from the Bray-curtis dissimilarity between successive inventories. Line colors represent the disturbance treatment (green for control, blue for T1,orange for T2 and red for T3).}\label{fig:NMDSplans}
\end{figure*}

@@

Except for leaf cholorphyll content, which continued to increase for
some T3 and T2 plots 30 years after disturbance, all traits and seed
mass proportions displayed a unimodal trajectories. The weighted means
of communities specific maximum height at adult stage (\emph{Hmax}),
leaf toughness (\emph{L\_toughness}) and wood specific gravity
(\emph{WD}) remained significantly lower than their initial value and
than these of the control plots (Figure \ref{fig:CWM}). The weighted
means of bark thickness (\emph{Bark\_thick}) similarly remained
substantilly higher than initially for all disturbed plots while the
specfic leaf area (\emph{SLA}) had almost recovered its initial value
and this of the undisturbed plots at the end of the experiment. Once
again the maximum difference to initial state was correlated to the
disturbance intensity
(\(\rho_{spearman}^{L_thickness}=0.67\),\(\rho_{spearman}^{L_chloro}=0.45\),\(\rho_{spearman}^{L_toughness}=-0.43\),\(\rho_{spearman}^{SLA}=0.93\),\(\rho_{spearman}^{WD}=-0.78\),\(\rho_{spearman}^{Bark-thickness}=0.88\),\(\rho_{spearman}^{Hmax}=-0.48\)).

@@

The functional diversity trajectories of plots based on the 8 leaf, stem
and life history traits was examined through the Rao diversity. For all
undisturbed plots the functional diversity remained comparable to the
reference values 5 years after disturbance (1989 inventories). For all
disturbed plots the functional trajectories followed a unimodal
trajectory with a maximum positively correlated to disturbance intensity
(\(\rho_{spearman}=XX\)) (Figure \ref{fig:DivFun}). The plot 7 from
treatment 1 that displayed a constantly outlying diversity was removed
from the graphical representation for better readability, see appendix
for full graph. Thirty years after disturbance all plots, whenever the
initial disturbance intensity, regained diversity values similar to
their initial avule and to those of control plots.

{\centering \includegraphics[width=0.8\linewidth]{WholePlotTrajectories_files/figure-latex/DivFun-1}

@@

\caption{Trajectories of the Rao functional diversity over 30 years after disturbance. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. Initial treatments are represented by solid lines colorswith green for control, blue for T1,orange for T2 and red for T3.and the missing trait value filling processes.}\label{fig:DivFun}
\end{figure*}

@@

The monitoring 30 years of disturbed forest communities confirmed the
limited direct impact of low disturbance on species richness as it was
found for impact survey of selective logging
\citep{Cannon1998, Baraloto2012a}. Only the most disturbed plots had not
fully achieved the ongoing recovery of genus richness, although all had
already reached equivalent levels as those of some control plots. Such
long time richness recovery had already been observed in several logging
experiments, which also highlighted the role of random disturbance
damage in the richness decrease \citep{DeAvila2015, Hu2018}.

The evenness of all disturbed plots and the richness of low disturbance
plots on the other hand mainly followed an asymptotic growth sharply
increasing until 15 years after disturbance. Such increased evenness is
explained by a much more homogeneous species distribution after
disturbance. Communities composition after disturbance is either due to
the old, pre-disturbance survivors or to the recruited trees. The
composition of old survivors proved to mirror the initial community
\citep{Herault2018}, the observed composition turnover would stem from
an enhanced the growth and survival of previously infrequent species
which, along with the IDH, reorganizes the typical high dominance
structure of hyperdiverse mature forests. This increase in taxonomic
diversity was accompanied by an increased taxonomic dissimilarity to
initial state along time, translating a a taxonomic turnover benefiting
to pioneers and light demanding species, as illustrated by the
communities functional shifts towards resource-acquisitive strategies
(sharp increase in the SLA, leaf thickness and bark thickness and
decrease in wood density, leaf toughness and maximum height)
\citep{Westoby1998, Wright2004, Reich2014}. Post-disturbance changes in
abiotic environment and competitive pressure then well favored pioneers
outcompeting other species when abiotic resources are abundant but
otherwise excluded in mature forests by long-lived, resistant and shade
tolerant species. The trajectories of proportions among seed mass
classes besides revealed the enhanced growth of small seeded, large
dispersive, species and therefore the importance of species reproductive
strategy which further supported the role of species dispersion and
demographic strategy for the post-disturbance dynamics, as suggested by
the IDH \citep{TerSteege2001, Flores2006, Haddad2008}.

@@

The functional diversity followed a hump-back trajectory which maximum
was positively correlated to the disturbance intensity
(\(\rho_{spearman}=\)) and after 30 years, all disturbed plots had
recovered diversities close to their initial values. Similar hump-back
trajectories were followed by all leaf and stem functional traits and
life history traits, to the exception of leaf chlorophyll content.
Although the return to initial values was still ongoing the recovery of
functional diversity and average communities trait values was
undeniable. This translated the recovery of ecosystems processes
\citep{Guariguata2001}, as functional traits are the most direct link
between biodiversity and ecosystem functioning \citep{Diaz2005}. Same
recovery trajectories were followed by taxonomic and functional
composition which argued for communities convergence and the maintenance
of plots initial difference
\citep{Hubbell1999, Molino2001, Baraloto2012a}. These compositon
difference, and probably also some abiotic parameters, besides strongly
determined the maximum and time path of the trajectories which, despite
an homogeneous trend were quite variable among plots.

@@

Despite the consistent recovery of initial diversity and composition
there was a time lag between the communities taxonomic and functional
characteristics. Taxonomic composition had longer time path and while
communities had recovered their functional diveristy, their taxonomic
composition and evenness remained higher than before disturbance. This
delay between functional and taxonomic dynamics was already observed for
grasslands \citep{Tilman1997, Mouillot2011} and more recently for
tropical forests \citep{Lohbeck2015, Guariguata2001}. Communities
functional diversity rely, according to the ``vegetation quantity
effect'' \citep{Grime1998}, on the functional strategy of dominant
species. Communities functional trajectory is then first driven by the
increase of dominant species diversity and evenness following
disturbance and then by the emergence of recruitment of species
progressively resembling the old pre-disturbance community which
decreased the diversity. Dominant species then restored the functional
diversity and the dominant functional type of the community although the
functional and taxonomic composition remained altered and the infrequent
species of initial communities are still missing to plots recovery. The
functional traits trajectories suggest that those species would match
the dominant functional type therefore increasing communities functional
redundancy. The functional overlap between species is typical of the
huge biodiversity of tropical forests \citep{Bellwood2006}, and proved
far from recovery 03 years after disturbance. It is though a determinant
of forests resilience and should be accounted for
\citep{Trenbath1999, Elmqvist2003, Diaz2005}.

Long-term alteration of functional redundancy besides came along with
persistent compositional changes which most probably impact disturbance
resistant species \citep{Haddad2008} and probably lianas or epiphytes
\citep{Martin2013}, and the modification of environmental conditions
like soils nutrient cycling and compaction \citep{Olander2005} forest
resilience remains under question \citep{Chazdon2003a}. New conditions
would not only be longer lasting but self-maintained as tied to
disturbance regime \citep{Burslem2000}. Specificaly, this would impair
species contingent to undisturbed forests, threatening their
maintenance, and run the risk to loose cornerstone species and trigger
unexpected ecological consequences
\citep{Jones1994, Diaz2005, Gardner2007}.

>>>>

In tropical forests communities are shaped by a constant range of
disturbance that changes the abiotic environment, as the light, heat and
water fluxes, and the biotic interaction and competitive pressure. The
cornerstone of tropical forests ecology is then to understand their
response to disturbance and its mechanisms
\citep{White2001, Chazdon2003a}. Forests response to disturbance has
been largely studied through structural parameters rapid and convenient
to measure, as aboveground biomass, tree height or stem density. These
structural parameters have then been sucessfully modeled and assessed
the maintenance of ecosystems processes and services
\citep{Denslow2000, Blanc2009, Rutishauser2016}. However the response of
tree species diversity although it is determinant of ecosystems
productivity, stability and functioning \citep{Tilman2014} and it would
be most probably impacted by the biotic and abiotic changes after
disturbance \citep{CazzollaGatti2014}.

@@

The taxonomic diversity was assessed through Richness and the Hill
number translation of Shannon and Simpson indices \citep{Hill1973}.

@@

\begin{figure}

@@

\caption{Trajectories of the plots in terms of \textbf{(a)} flora composition and \textbf{(b)} functional composition regarding the 6 leaf and stem functional traits,the maximum allometric height and seed mass class in the two-dimensional space from the NMDS performed for the 30 years after disturbance. Distance matrix for NMDS were computed from the Bray-curtis dissimilarity between successive inventories. Line colors represent the disturbance treatment (green for control, blue for T1,orange for T2 and red for T3).}\label{fig:NMDSplans}
\end{figure}

@@

Except for leaf cholorphyll content, which continued to increase for
some T3 and T2 plots 30 years after disturbance, all traits and seed
mass proportions displayed a unimodal trajectories. The weighted means
of communities specific maximum height at adult stage (\emph{Hmax}),
leaf toughness (\emph{L\_toughness}) and wood specific gravity
(\emph{WD}) remained significantly lower than their initial value and
than these of the control plots (Figure \ref{fig:CWM}). The weighted
means of bark thickness (\emph{Bark\_thick}) similarly remained
substantilly higher than initially for all disturbed plots while the
specfic leaf area (\emph{SLA}) had almost recovered its initial value
and this of the undisturbed plots at the end of the experiment. Once
again the maximum difference to initial state was correlated to the
disturbance intensity (\(\rho_{spearman}^{L_thickness}=0.67\),
\(\rho_{spearman}^{L_chloro}=0.45\),
\(\rho_{spearman}^{L_toughness}=-0.43\), \(\rho_{spearman}^{SLA}=0.93\),
\(\rho_{spearman}^{WD}=-0.78\),
\(\rho_{spearman}^{Bark-thickness}=0.88\),
\(\rho_{spearman}^{Hmax}=-0.48\)).

@@

The functional diversity trajectories of plots based on the 8 leaf, stem
and life history traits was examined through the Rao diversity. For all
undisturbed plots the functional diversity remained comparable to the
reference values 5 years after disturbance (1989 inventories). For all
disturbed plots the functional trajectories followed a unimodal
trajectory with a maximum positively correlated to disturbance intensity
(\(\rho_{spearman}=0.73\)) (Figure \ref{fig:DivFun}). The plot 7 from
treatment 1 that displayed a constantly outlying diversity was removed
from the graphical representation for better readability, see appendix
for full graph. Thirty years after disturbance all plots, whenever the
initial disturbance intensity, regained diversity values similar to
their initial avule and to those of control plots.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/DivFun-1}

@@

\caption{Trajectories of the Rao functional diversity over 30 years after disturbance. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. Initial treatments are represented by solid lines colorswith green for control, blue for T1,orange for T2 and red for T3.and the missing trait value filling processes.}\label{fig:DivFun}
\end{figure}

@@

The monitoring of disturbed forest communities confirmed a limited
negative impact of disturbance on species richness, as observed on
several post logging surveys \citep{Cannon1998, Baraloto2012a}. Thirty
years after disturbance, the genus richness was restored to initial plot
values after high disturbance intensity while it substantially increased
after low disturbance intensity, reaching a gain of almost 12 genuses
for some plots.

Both richness and evenness after disturbance followed asymptotic
trajectories, sharply increasing for 15 years before stabilizing at
higher values than those of control plots. An increase of communities
evenness stems from a higher homogeneity of species distribution. After
disturbance, communities are made of the old, pre-disturbance survivors
and the newly recruited trees: changes in composition and abundance
distribution are to be found in the recruitment processes and in the
pre-disturbance survivors mortality. The composition of old survivors
proved to mirror the initial community \citep{Herault2018}, which leads
to support a species turnover among recruited trees with enhanced growth
and survival of previously infrequent species. Indeed, the increase in
taxonomic diversity was accompanied by an increase of taxonomic
dissimilarity with plots initial state and a functional shift towards
resource-acquisitive strategies (sharp increase in the SLA, leaf
thickness and bark thickness and decrease in wood density, leaf
toughness and maximum height)
\citep{Westoby1998, Wright2004, Reich2014}. Disturbance then causes a
reorganization of the typical high dominance structure of hyperdiverse
mature forests after disturbance, benefiting to pioneers and light
demanding species. Likely, the changes in abiotic environment and
competitive pressure favored pioneers which outcompete others in non
limiting resources but are excluded in mature forests by long-lived,
more resistant and shade tolerant species.

As stated by the IDH, communities dynamics after disturbance then rely
on species functional strategy and corresponding ability to fill the
environmental niches made available by disturbance. Recruited species
then mix with pre-disturbance ones from which they differ and constitue
a community all the more diversified that the disturbance is intense
\citep{Molino2001}.

@@

The trajectories of almost all leaf and stem functional traits, life
history traits,and functional diversity followed a hump-back trajectory
which maximum were positively correlated to the disturbance intensity,
and were recovering towards initial values after 30 years that
trajectory. Functoinal traits and communities functional diversity are
the most direct link between biodiversity and ecosystem functioning
\citep{Diaz2005}: these trajectories reflected a consistent recovery in
the long term of ecosystem processes \citep{Guariguata2001}.

Same hump-back trajectories were followed by the composition of plots
species and functoinal composition. This similarly meant a recovery of
the pre-disturbance state and the maintenance of the initial differences
among plots which supports the idea of a convergence of communities
composition \citep{Hubbell1999, Molino2001, Baraloto2012a}. Although
plots followed homogeneous trends these initial differences in
compositon difference, and probably also some abiotic parameters,
strongly determined the maximum and time path of the trajectories
\citep{Anderson2007}•.

@@

Despite the consistent recovery of initial diversity and composition
there was a time lag between the trajectories followed by taxonomic and
functional characteristics. Taxonomic composition had longer time path
and while communities had recovered their functional characteristics,
their taxonomic composition and evenness remained altered by
disturbance. This delay between functional and taxonomic dynamics was
already observed for grasslands \citep{Tilman1997, Mouillot2011} and
more recently for tropical forests \citep{Lohbeck2015, Guariguata2001}.
The functional diversity of communities rely on the functional strategy
of dominant species, according to the ``vegetation quantity effect''
\citep{Grime1998}. The functional trajectories of communities are then
driven first by the increase of diversity and evenness of dominant
species and then by the recruitment of species resembling the old
pre-disturbance community which reduced the overall diversity. Although
dominant species restoredthe functional pre-disturbance characteristics,
communities composition remained altered and the infrequent species of
initial communities remain missing to achieve plots recovery. According
to the funcitonal traits trajectories, those infrequent species would
match the dominant functional type and would participate to the
functional redundancy. The functional redundancy, thas is the functional
overlap between species that is typical of the huge biodiversity of
tropical forests \citep{Bellwood2006}, therefore was not restored 30
years after disturbance and this is majr to consider as it defines
forests' resilience \citep{Trenbath1999, Elmqvist2003, Diaz2005}.

Besides the long-term alteration of functional redundancy, the
persistent compositional changes which might have favored disturbance
resistant species \citep{Haddad2008}, lianas or epiphytes
\citep{Martin2013} and would probably have come along with environmental
change, like in the soils nutrient cycling and compaction
\citep{Olander2005}. These presistent changes highly question forest's
resilience\citep{Chazdon2003a}. New conditions would not only be longer
lasting but self-maintained as tied to disturbance regime
\citep{Burslem2000}. Specificaly, this would impair species contingent
to undisturbed forests, threatening their maintenance, and run the risk
to loose cornerstone species and trigger unexpected ecological
consequences \citep{Jones1994, Diaz2005, Gardner2007}.
